Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-22 final_metrics_analysis.Rmd
changing color on mosaic plot, reordering graphs

```{python mosiac of charge degrees by race}
props={}
props[('Caucasian','F')]={'facecolor':'red', 'edgecolor':'white'}
props[('Caucasian','M')]={'facecolor':'red', 'edgecolor':'white'}
props[('African-American','F')]={'facecolor':'xkcd:aqua','edgecolor':'white'}
props[('African-American','M')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
labelizer=lambda k:{('Caucasian','F'):1242,('African-American','F'):2194,('Caucasian','M'):858,('African-American','M'):979}[k]
mosaic(df_viz,['race','c_charge_degree'],labelizer=labelizer,properties=props)
plt.show()
```

59.14% of white defendants were charged with felonies versus 69.15% of black defendants. Likewise, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).

@@

```{python exploratory recidivism mosaic plot}
props={}
props[('Caucasian','Recidivated')]={'facecolor':'green', 'edgecolor':'white'}
props[('Caucasian','Survived')]={'facecolor':'green', 'edgecolor':'white'}
props[('African-American','Recidivated')]={'facecolor':'xkcd:purple','edgecolor':'white'}
props[('African-American','Survived')]=        {'facecolor':'xkcd:purple','edgecolor':'white'}
labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
plt.show()
```

@@

@@

In the former plot, we can see that Black defendants have a higher number of prior charges than do white defendants. This again likely reflects the over-policing of Black people and communities. In the latter plot, we can see that male defendants have a higher number of prior charges than do female defendants. This is worth noting as @larson2016we found that there is discrepancy in COMPAS risk assessment algorithm's prediction between sexes. When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes [@larson2016we].

>>>>

```{python mosiac of charge degrees by race}
props={}
props[('Caucasian','F')]={'facecolor': green, 'edgecolor':'white'}
props[('Caucasian','M')]={'facecolor':green, 'edgecolor':'white'}
props[('African-American','F')]={'facecolor':purple,'edgecolor':'white'}
props[('African-American','M')]=        {'facecolor':purple,'edgecolor':'white'}
labelizer=lambda k:{('Caucasian','F'):1242,('African-American','F'):2194,('Caucasian','M'):858,('African-American','M'):979}[k]
mosaic(df_viz,['race','c_charge_degree'],labelizer=labelizer,properties=props)
plt.show()
```

1242 (59.14%) of white defendants were charged with felonies versus 2194 (69.15%) of Black defendants. Additionally, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).

@@

@@

```{python exploratory recidivism mosaic plot}
props={}
props[('Caucasian','Recidivated')]={'facecolor':green, 'edgecolor':'white'}
props[('Caucasian','Survived')]={'facecolor':green, 'edgecolor':'white'}
props[('African-American','Recidivated')]={'facecolor':purple,'edgecolor':'white'}
props[('African-American','Survived')]=        {'facecolor':purple,'edgecolor':'white'}
labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
plt.show()
```

@@

In the above plot, we can see that Black defendants, particularly men, are more likely to have a greater count of prior charges than white defendants. This again likely reflects the over-policing of Black people and communities. Additionally, we can see that male defendants have a higher number of prior charges than do female defendants. This is worth noting as ProPublica found that there is discrepancy in COMPAS risk assessment algorithm's prediction between sexes. When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes (ProPublica). Though we do not know for sure which information goes into the COMPAS algorithm, it is likely that a defendant with prior charges will be coded as a having a higher risk of recidivism. Thus, by looking at the racial discrepancies in prior charges we can already see potential bias in the algorithm.
